Classless Java: Tuning Java Interfaces

### 2015-12-02 paper/sections/Formalization.tex
Revise translation with @ObjWeak added.
revising text

\section{Interaction of interface methods with interface composition}
From Java8 interfaces can have three type of methods:
abstract methods, default methods, static methods.
\begin{itemize}
\item Static methods are handled in a very clean way: they are visible only on the interface they are explicitly declared. This means that 
the following code is ill-typed.
\begin{lstlisting}
interface A0 { static int m(){return 1;} }
interface B0 extends A0 {}
...
B0.m()//ill typed
\end{lstlisting}
Note how this is different w.r.t. the way static methods are handled in classes.
In this way static methods have simply no interaction with interface composition (\Q@extends@ or \Q@implements@).

\item Abstract methods composition is accepted when there is one that is the most specific.  For example in 
method \texttt{Integer m()} is visible in \Q@C1@.

\begin{lstlisting}
interface A1 { Object m(); }
interface B1 { Integer m();}
interface C1 extends A1,B1 {} //accepted
\end{lstlisting}

\item Default methods conflict with any other default or abstract method. For
example the following code is rejected due to conflicting methods.
\begin{lstlisting}
interface A2 { default int m() {return 1;}}
interface B2 { int m(); }
interface C2 { default int m() {return 2;}}
interface D2 extends A2,B2 {} //rejected due to conflicting methods
interface E2 extends A2,C2 {} //rejected due to conflicting methods
\end{lstlisting}
Note how this is in contrast with what happens in most trait models, where \Q@D2@ would be accepted, and the implementation in \Q@A2@ would be part of the behaviour of \Q@D2@.

\item The method in the current interface wins over any methods defined in its
super-interfaces, provided that the method conform to the subtype of all methods
in its super-interfaces, i.e., the method is the most specific one.
This also override rejection due to conflicting methods. For example, the following code is accepted, but would be rejected (see before) if the method \Q@m@ was not redefined.
\begin{lstlisting}
interface D3 extends A2,B2 { int m(); } //accepted
interface E3 extends A2,C2 { default int m(){return 42;} } //accepted
\end{lstlisting}
\end{itemize}

@@

What happens if we define a new interface \Q@M@ extending one \Q@A@${}_i$ and one \Q@B@${}_i$? we have 9 cases, that can fit nicely a table:\\*
\noindent
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & conservative error                        & conflict error      & conservative error       \\ \hline
\textbf{B2}        & both abstract, accepted                        & conservative error       & both abstract, accepted       \\ \hline
\textbf{B3}        & \textbf{conservative error} &conflict  error       & conservative error      \\ \hline
\end{tabular}
\\*
We try to classify the results out of the table:
\begin{itemize}
\item \textbf{conflict error} happens when the method from \Q@A@${}_i$ and one \Q@B@${}_i$ are both implemented. This is also considered an error in most trait models.
\item \textbf{both abstract, accepted} happens when the method from \Q@A@${}_i$ and one \Q@B@${}_i$ are both abstract. This is also considered correct in all trait models.
\item \textbf{conservative error}
 happens when the method from \Q@A@${}_i$ and one \Q@B@${}_i$ is implemented in only one side.
This is different from what we would expect in a trait model, but is coherent with the conservative idea that a method defined in an interface
should not silently satisfy a method in another one.
\end{itemize}

During our experimentation, we found a bug in ECJ (eclipse version of javac): The case \textbf{B3},\textbf{A1}
is accepted by ECJ4.5.1 and rejected by javac.
By email communication with Brian Goetz (lead Java8 designer) we have confirmation that the expected behaviour is rejection, and that this is a bug in ECJ.
%Table~\ref{table:javabug} shows the method overriding bug in Java. In the
%example, there are 6 interfaces \texttt{A1,A2,A3,B1,B2,B3} with methods all
%named as \texttt{m} which are either abstract or default methods. If we define a
%new interface \texttt{M} that extends two of these interfaces, then the method
%overriding result is shown in the table. For example, row2 col2 means \texttt{M
%  extends A1,B1}. The result is \texttt{ERROR} because the abstract method
%\texttt{T m();} in \texttt{A1} conflicts with the method \texttt{default T m()
%  {...}} in \texttt{B1}. Readers may also figure out other extending cases by
%following this way of interpretation.

@@

\begin{figure}[t]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of ClassLess Java}
\label{Grammar}
\end{figure}

>>>>

\section{Interaction of Interface Methods with Interface Composition}
From Java 8 interfaces can have three type of methods: abstract, default, and
static methods.
\begin{itemize}

\item Static methods are handled in a very clean way: they are visible only in
  the interface they are explicitly defined. This means the following code
  is ill-typed.
\begin{lstlisting}
interface A0 { static int m(){return 1;} }
interface B0 extends A0 {}
... B0.m()//ill typed
\end{lstlisting}
This is different from the way static methods are handled in classes. Here
static methods have simply no interaction with interface
composition (\Q@extends@ or \Q@implements@).

\item Abstract methods composition is accepted when exist a most specific one.
  For example, here method \texttt{Integer m()} from \texttt{B1} is visible in \Q@C1@.
\begin{lstlisting}
interface A1 { Object m(); }
interface B1 { Integer m();}
interface C1 extends A1,B1 {} //accepted
\end{lstlisting}

\item Default methods conflict with any other default or abstract method. For
  example the following code is rejected due to methods conflict.
\begin{lstlisting}
interface A2 { default int m() {return 1;}}
interface B2 { int m(); }
interface C2 { default int m() {return 2;}}
interface D2 extends A2,B2 {} //rejected due to conflicting methods
interface E2 extends A2,C2 {} //rejected due to conflicting methods
\end{lstlisting}
Note how this is different from what happens in most trait models, where \Q@D2@
would be accepted, and the implementation in \Q@A2@ would be part of the
behaviour of \Q@D2@.

\item A method in the current interface wins over any method in its
super-interfaces, provided that the method%  conforms to the subtype of all methods
% in its super-interfaces, i.e., the method
is the most specific one. This method also overrides conflict due to
inheritance. For example, the following code is accepted, but would be rejected
(see before) if the method \Q@m@ was not redefined in \texttt{D3} and
\texttt{E3}.
\begin{lstlisting}
interface D3 extends A2,B2 { int m(); } //accepted
interface E3 extends A2,C2 { default int m(){return 42;} } //accepted
\end{lstlisting}

\end{itemize}

@@

What happens if a new interface \Q@M@ extends one \Q@A@${}_i$ and one
\Q@B@${}_i$? The 9 cases are shown here:\\*
\noindent
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{M extends} & \textbf{A1}                  & \textbf{A2} & \textbf{A3} \\ \hline
\textbf{B1}        & conservative error                        & conflict error      & conservative error       \\ \hline
\textbf{B2}        & both abstract, accepted                        & conservative error       & both abstract, accepted       \\ \hline
\textbf{B3}        & \textbf{conservative error} &conflict  error       & conservative error      \\ \hline
\end{tabular}
\\*
We try to classify the results out of the table:
\begin{itemize}
\item \textbf{conflict error} happens when the methods from \Q@A@${}_i$ and
  \Q@B@${}_i$ are both implemented, which is also an error in most trait models.
\item \textbf{both abstract, accepted} happens when the methods from \Q@A@${}_i$
  and \Q@B@${}_i$ are both abstract, which is also considered correct in all
  trait models.
\item \textbf{conservative error} happens when only one method from \Q@A@${}_i$
  or \Q@B@${}_i$ is implemented, which is different from what we would expect in
  a trait model, but is coherent with the conservative idea that a method
  defined in an interface should not silently satisfy a method in another one.
\end{itemize}

During our experimentation, we found a bug in ECJ (Eclipse compiler for Java):
the case \textbf{M} extending \textbf{B3} and \textbf{A1} is accepted by
ECJ4.5.1 and rejected by javac.  By email communication with Brian Goetz
(leading Java 8 designer) we have confirmed that the expected behaviour is
rejection, hence this is a bug in ECJ.\footnote{This bug is fixed in the ECJ
  developer brunch , but not released as a stable version yet.}
%Table~\ref{table:javabug} shows the method overriding bug in Java. In the
%example, there are 6 interfaces \texttt{A1,A2,A3,B1,B2,B3} with methods all
%named as \texttt{m} which are either abstract or default methods. If we define a
%new interface \texttt{M} that extends two of these interfaces, then the method
%overriding result is shown in the table. For example, row2 col2 means \texttt{M
%  extends A1,B1}. The result is \texttt{ERROR} because the abstract method
%\texttt{T m();} in \texttt{A1} conflicts with the method \texttt{default T m()
%  {...}} in \texttt{B1}. Readers may also figure out other extending cases by
%following this way of interpretation.

@@

\begin{figure}[t]
\begin{grammar}
\production{
\e
}{
  \x\mid\MCall\e\m\es\mid\MCall{\C}\m\es\mid\MCall{\C\QM{.super}}\m\es\mid\x\QM=\e\QM;\e'\mid\obj
  }{expressions}\\
\production{
\obj
}{
\QM{new}\ \C\oR\cR\oC\fields\
\mh_1\oC\QM{return}\ \e_1\QM{;}\!\cC
\ldots
\mh_n\oC\QM{return}\ \e_n\QM{;}\!\cC
\cC
  }{object creation}\\
\production{\field}{\T\ \f \QM= \x\QM;}{field declaration}\\
\production{
\II
}{
 \ann\ \QM{interface}\ \C\ \QM{extends}\ \Cs\ \oC \methods\ \cC
  }{interface declaration}\\
\production{
\method
}{
 \QM{static}\ \mh\ \oC\QM{return}\ \e\QM{;}\!\cC
\mid
\QM{default}\ \mh\ \oC\QM{return}\ \e\QM{;}\cC
\mid
\mh\QM{;}
  }{method declaration}\\
\production{
\mh
}{
 \T_0\ \m\ \oR\T_1\ \x_1\ldots\T_n\ \x_n\cR
  }{method header}\\
\production{
\ann
}{
  \mixinAnn|\weakAnn|\emptyset
  }{annotations}\\
\production{\Gamma}{
\x_1{:}\C_1\ldots\x_n{:}\C_n
}{environment}
\end{grammar}
\caption{Grammar of ClassLess Java}
\label{Grammar}
\end{figure}

### 2015-12-02 paper/sections/Overview.tex
revising text

\subsection{Advanced Multiple Inheritance with Object Interfaces}
Finally, we can define Pegasus as simply as we did in the simplified (and stateless) 
version in the right-side of Figure~\ref{fig:pegasus}:

@@

\noindent Note how even the non trivial pattern for field type refinement is
transparently composed, and \texttt{Pegasus} has a \Q@Point3D@
\Q@location@. This works because \Q@Horse@ do not perform
any field type refinement, otherwise we may have to choose/create a
common subtype in order for \texttt{Pegasus} to exists.

>>>>

\subsection{Advanced Multiple Inheritance with Object Interfaces}
Finally, we can define \texttt{Pegasus} as simple as we did in the simplified
(and stateless) version on the right-side of Figure~\ref{fig:pegasus}:

@@

\noindent Note how even the non trivial pattern for field type refinement is
transparently composed, and \texttt{Pegasus} has a \Q@Point3D@
\Q@location@.%  This works because \Q@Horse@ do not perform
% any field type refinement, otherwise we may have to choose/create a
% common subtype in order for \texttt{Pegasus} to exists.
